---
title: "Feature extraction and interpretation from convolutional neural networks"
author: "Raphael Mourad, Maître de Conférences, Université Paul Sabatier"
date: "10/02/2021"
output: html_document
---

## SETUP PROJECT FOLDER

Set up your project folder to your folder "atelier_INSERM".
```{r, setup}
knitr::opts_knit$set(root.dir = "/home/mourad/Downloads/Deep-Learning-for-Genomics-Training-main/")
knitr::opts_chunk$set(echo = TRUE)
```

## AIM OF THE SCRIPT

This script is used to extract and interpret the convolutional features from the trained CNN.


## LOAD LIBRARIES


Load R librairies.
Remove suppressPackageStartupMessages() for debugging!
```{r, libraries}
suppressPackageStartupMessages(library(GenomicRanges))
suppressPackageStartupMessages(library(Biostrings))
suppressPackageStartupMessages(library(TFBSTools))
suppressPackageStartupMessages(library(JASPAR2020))
#suppressPackageStartupMessages(library(reticulate))
suppressPackageStartupMessages(library(motifStack))
suppressPackageStartupMessages(library(universalmotif))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(grImport2))
```

## LOAD FUNCTIONS

```{r, function}
suppressPackageStartupMessages(source("scriptR/functions.R"))
```

## SOME PARAMETERS

Define parameters for the analysis:  
- peakSize is the resized peak size. For machine/deep learning, it is easier to work with peaks have the same size.  
- kpeaks is the number of best peaks used. Note that all peaks can also be used.  
- expe is the ChIP-seq experiment. For instance, CTCF for CTCF ChIP-seq or POL2 for RNA Polymerase 2 ChIP-seq experiment. 
```{r, parameters}
peakSize=201
kpeaks=4000
expe="DNase"
DNAletters=c("A","C","G","T")
vocab_size=length(DNAletters)
```

## INPUT FILE NAMES

Set file names, which includes:  
- positive peaks from ChIP-seq and negative peaks from the control generator function genNullSeqs().  
- bed (genomic ranges) and fasta files (corresponding DNA sequences).
```{r, files}
fileBedPos=paste0("data/bed/",expe,"_GM12878_hg19_",kpeaks,"_pos.bed")
fileBedNeg=paste0("data/bed/",expe,"_GM12878_hg19_",kpeaks,"_neg.bed")
fileFastaPos=paste0("data/fasta/",expe,"_GM12878_hg19_",kpeaks,"_pos.fa")
fileFastaNeg=paste0("data/fasta/",expe,"_GM12878_hg19_",kpeaks,"_neg.fa")
```

## LOAD PROCESSED DATA

Load ChIP-seq and control fasta sequences:
```{r, read}
peakPos.seq=readDNAStringSet(fileFastaPos)
peakNeg.seq=readDNAStringSet(fileFastaNeg)

print(peakPos.seq)
```

## FEATURE EXTRACTION

Load conv activation values from Google Colab
```{r}
data_conv=fread("../conv_activations_pos_2d.csv.gz",header=F)
data_conv=as.matrix(data_conv)
print(dim(data_conv))

conv1d_output <- array(data_conv, dim = c(nrow(data_conv), 178, 16))
dim(conv1d_output)

print(conv1d_output[1,1:10,1])
```


Convert kernels to Position Frequency Matrices (PFMs)
```{r, convert_kernels}
kernelSize=dim(conv1d_output)[3]
activationThreshold=0.15
kernelPFMList=convertToPWMs(conv1d_output, peakPos.seq, activationThreshold)
print(kernelPFMList)
```

Plot first motif
```{r, plot_1stmotif}
motif1 <- new("pcm", mat=as.matrix(kernelPFMList[[2]]), name=ID(kernelPFMList[[2]]))
plot(motif1)
```

Trimming and filtering of motifs based on information content
```{r, trim_filter}
ICthreshold=0.7
PFMTrimmedList=motifTrimming(kernelPFMList,ICthreshold)
motifTrimmed1 <- new("pcm", mat=as.matrix(PFMTrimmedList[[1]]), name=ID(PFMTrimmedList[[1]]))
plot(motifTrimmed1)
```

Export motifs for dimension reduction with RSAT matrix-clustering
```{r, export_motifs}
kernelPFMListUM=convert_motifs(PFMTrimmedList, class = "universalmotif-universalmotif")
write_jaspar(kernelPFMListUM,paste0("results/motif/trimmed_redundant_motifs_",expe,".jaspar"),overwrite=TRUE)
```


## MOTIF DIMENSION REDUCTION

On https://rsat01.biologie.ens.fr/rsat/matrix-clustering_form.cgi  
Run matrix clustering using the transfac file of motifs.  
After, download the root motifs (click on "Additional Files").  
Rename it to paste0("root_motifs_",expe,".tf").  


## FEATURE IMPORTANCE

Load root motifs
```{r, load_root_motifs}
rootMotifsUM=read_transfac(paste0("results/motif/root_motifs_",expe,".tf"))
rootMotifs=do.call(PFMatrixList,convert_motifs(rootMotifsUM, "TFBSTools-PFMatrix"))
print(rootMotifs)
```

DNA motif counts
```{r, motif_counts}
# Bind positive and negative sequences, and make label
peakAll.seq=c(peakPos.seq,peakNeg.seq)
label=c(rep(1,length(peakPos.seq)),rep(0,length(peakNeg.seq)))


# Shuffle sequence indices
idxS=sample(1:length(peakAll.seq))
peakAllS.seq=peakAll.seq[idxS]
labelS=label[idxS]

motif_ix=matchMotifs(rootMotifs,peakAllS.seq,out="scores",p.cutoff=1e-4)
mcAllS=motifCounts(motif_ix)


# Split train and test indices
percTrain=0.7
idxTrain=1:(ceiling(length(labelS)*percTrain))
idxTest=(length(idxTrain)+1):length(labelS)
labelTrain=labelS[idxTrain]
labelTest=labelS[idxTest]

mcTrain=mcAllS[idxTrain,]
mcTest=mcAllS[idxTest,]
print(mcTrain[1:5,])
```

Random forests
```{r, randomforests}
dataRF_motif=data.frame(label=labelTrain,as(mcTrain,"matrix"))
RF_motif=ranger(label ~ .,data=dataRF_motif,importance="permutation")
predRF_motif=predict(RF_motif,data=data.frame(as(mcTest,"matrix")))$predictions
rocRF_motif=pROC::roc(as.factor(labelTest),predRF_motif,ci=T)
plot(rocRF_motif,main=paste0("AUROC: ", round(pROC::auc(rocRF_motif),3)))
```

Variable importance
```{r, varimp}
motifID=ID(rootMotifs)
dataImportanceMotif=data.frame(motifID,importance=importance(RF_motif))
dataImportanceMotif=dataImportanceMotif[order(importance(RF_motif),decreasing=T)[1:20],]
dataImportanceMotif2=dataImportanceMotif[order(dataImportanceMotif[,2]),]
barplot(dataImportanceMotif2[,2],names.arg=dataImportanceMotif2[,1],cex.names=0.7,horiz=T,las=2,xlab="Importance")
```

## FEATURE (MOTIF) VISUALIZATION

Plot best motif logo from PFM matrix
```{r, plot_bestmotif}
rootMotifsMS=lapply(rootMotifs,function(x){new("pcm", mat=as.matrix(x), name=ID(x))})
idxBestRootMotif=which(ID(rootMotifs)==dataImportanceMotif[1,1])
plot(rootMotifsMS[[idxBestRootMotif]])
```

Plot motif clustering tree
```{r, plot_clustering, fig.width=14, fig.height=26}
file_plot_tree_motifs=paste0("results/motif/plot_tree_motifs_",expe,".pdf")
#pdf(file_plot_tree_motifs,30,200)
motifStack(rootMotifsMS, layout="tree")
#dev.off()
```









